---
title: "Cascaded experiment"
author: "Manuel Rivera"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Description
One approach to generate keystone datasets controls is the use of cascaded interactions matrixes. Suppose a 5 species community, if we set the diagonal to a certain value and we can craft a cascading effect on the matrix.
A represation of this process is shown below, setting keystone specie to 1, we can generate a matrix with the next structure:
Specie 1 -> Specie 2 -> Specie 3 -> Specie 4 -> Specie 5
Or peerhaps a more complex one:
Specie 1 -> Specie 5 -> Specie 4 -> Specie 2 -> Specie 3

For testing we will run the metric distributions:

# Fixing needed
What if boosted interaction is 0
# Generate data
```{r, eval=FALSE}

dir = '/mnt/data/sur/users/mrivera/Controls/exp_20251125'
simulation_parameters = data.table::fread(paste0(dir, "/Simulation-parameters.tsv"), select = c("id", "keys", "p_noint"))
total_rows = nrow(simulation_parameters) * 30

ext_dir = '/mnt/data/sur/users/mrivera/Controls/exp_20251125/GNN-targets'
raw_outs_dir = '/mnt/data/sur/users/mrivera/Controls/exp_20251125/raw-ODEs'

library(dplyr)
library(parallel)

process_row <- function(i) {
  #------------------------
  # Generate path
  row <- simulation_parameters[i, ]
  id <- row$id
  key <- as.numeric(row$key)
  p_noint <- as.numeric(row$p_noint)
  ext_file_path <- paste0(ext_dir, paste0("/tgt_", id, ".feather"))
  ode_file_path <- paste0(raw_outs_dir, paste0("/O_", id, ".feather"))
  #------------------------
  # Generate relative abundance
  population <- arrow::read_feather(ode_file_path, col_select = 1000)
  relative_pop <- population / sum(population)
  #------------------------
  # Read extinctions file
  cols <- c("new_ext", "BC_diss", "K_s")
  table <- arrow::read_feather(ext_file_path, col_select = all_of(cols)) %>%
    rename(n_extinctions = new_ext) %>%
    rename(dissimilarity = BC_diss) %>%
    rename(keystoneness = K_s) %>%
    mutate(pop_initial = population[[1]]) %>%
    mutate(rel_pop_initial = relative_pop[[1]]) %>%
    mutate(prop_extinctions = n_extinctions / n()) %>%
    mutate(is_keystone = FALSE, p_noint = p_noint)
  # Add keystone
  table$is_keystone[key] <- TRUE
  table
}

results <- mclapply(1:nrow(simulation_parameters), process_row, mc.cores = detectCores() - 1)
full_summary <- do.call(rbind, results)

#-----------------------
# Section: Convert to long format.
library(tidyr)
full_summary_long <- full_summary %>%
    pivot_longer(
        cols = c(keystoneness, dissimilarity, n_extinctions, pop_initial, prop_extinctions, rel_pop_initial),
        names_to = "variable",
        values_to = "value"
    ) %>%
    dplyr::mutate(value = replace_na(value, 0))


#-----------------------
# Section: Generate directory for plots
plt_dir = paste0('/mnt/data/sur/users/mrivera/Plots/', basename(dir))
if (!(dir.exists(plt_dir))) { 
  dir.create(plt_dir)
  print('>> Directory created\n')}
```

# Metric distribution
```{r, eval=FALSE}

```
