---
title: "Cascade experiment"
author: "Manuel Rivera"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Description
This analysis examines the cascade approach for generating keystone dataset controls. In this method, off-diagonal 
elements are set to zero or negative by default, with the exception of a select few that are assigned a negative interaction strength drawn from `-U(0,1)*10`.
To illustrate, consider a community of 5 species where species 3 is the keystone species. The resulting interaction matrix would take the following form:
Interactions would follow the pattern: Specie 3 -> Specie 1, Specie 3 -> Specie 2, Specie 3 -> Specie 4, and Specie 3 -> Specie 5. 
The remaining interactions would be set to zero or negative values.
```{r, eval=TRUE}
# Generate a 5x5 matrix 
n_species <- 5
k = 3 # keystone species index
num_noint = 10 # number of zero interactions
#--------------------
# Generate matrices
mask <- matrix(TRUE, n_species, n_species)    # mask matrix
M = matrix(NA, n_species, n_species)          # initialize interaction matrix
diag(mask) <- FALSE                           # remove diagonal
whom_rows = seq_len(n_species)[-k] # species affected by the cascade effect
who_cols = c(k, whom_rows)     # who
#-------------------
# Mark positions affected by the cascade effect
for (i in seq_along(whom_rows)) {
  # Update the mask 
  mask[whom_rows[i], who_cols[i]] = FALSE
  # Fill M matrix
  M[whom_rows[i], who_cols[i]] = -runif(1, min = 0, max = 1)  * 10 
  # Print order
  cat("Specie", whom_rows[i], "(row) affected by", who_cols[i], "(column)\n", sep = " ")
}
#------------------
# Fill diagonal
diag(M) <- -0.5
num_negs = sum(mask) - num_noint # number of negative interactions
M[mask] <- c(rep(0, num_noint),-runif(num_negs, min = 0, max = 1))
print(M)
```

# Generate data
```{r, eval=FALSE}

dir = '/mnt/data/sur/users/mrivera/Controls/Cascade_keystone'
ext_dir = paste0(dir, "/ExtSummaries")
raw_outs_dir = paste0(dir, "/RawOutputs")
simulation_parameters = data.table::fread(paste0(dir, "/simulation-params.tsv"), select = c("id", "key", "p_noint"))
i = 1

library(dplyr)
library(parallel)

process_row <- function(i) {
  #------------------------
  # Generate path
  row <- simulation_parameters[i, ]
  id <- row$id
  key <- as.numeric(row$key)
  p_noint <- as.numeric(row$p_noint)
  ext_file_path <- paste0(ext_dir, paste0("/ExtSummary_", id, ".feather"))
  #------------------------
  # Read extinctions file
  cols <- c("prop_extinctions", "dissimilarity_bc", "keystoneness", "rel_pop_initial")
  table <- arrow::read_feather(ext_file_path, col_select = all_of(cols)) %>%
    rename(dissimilarity = dissimilarity_bc) %>%
    rename(keystoneness = keystoneness) %>%
    #mutate(pop_initial = population[[1]]) %>%
    mutate(rel_pop_initial = rel_pop_initial) %>%
    mutate(prop_extinctions = prop_extinctions) %>%
    mutate(is_keystone = FALSE, p_noint = p_noint) %>%
    mutate(specie = 1:n()) %>%
    mutate(id = id)
  # Add keystone
  table$is_keystone[key] <- TRUE
  table
}

results <- mclapply(1:nrow(simulation_parameters), process_row, mc.cores = detectCores() - 1)
full_summary <- do.call(rbind, results)
```

# Distributions metrics
```{r, eval=FALSE}
library(ggplot2)
library(ggridges)

# Define common theme and colors
theme_custom <- theme_minimal(base_size = 12) +
    theme(
        legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 11),
        axis.title = element_text(size = 10)
    )

colors <- c("FALSE" = "#4575b4", "TRUE" = "#d73027")  # Better color contrast

# Only plot these variables
variables_to_plot <- c("keystoneness", "dissimilarity", "prop_extinctions")

# Generate plot
library(tidyr)
plt = full_summary %>% select(is_keystone, all_of(variables_to_plot)) %>%
    pivot_longer(cols = all_of(variables_to_plot), names_to = "variable", values_to = "value") %>%
    mutate(value = ifelse(value < 0, 0, value)) %>%
    ggplot(aes(x = variable, y = value, fill = is_keystone)) +
    geom_boxplot(alpha = 0.6, outlier.shape = NA) +
    scale_y_continuous(trans = "log1p") +
    stat_summary(fun = mean, geom = "point", color = "red", size = 3, position = position_dodge(width = 0.75), show.legend = FALSE) +
    scale_fill_manual(values = colors, labels = c("Non-keystone", "Keystone")) +
    labs(x = "Metric", y = "Value", fill = "Species Type") +
    theme_custom

#------------------------------
# Section: Save plot
path = file.path("/mnt/data/sur/users/mrivera/Plots/Cascade_keystone", "Metric_distributions.png")
if (!(dir.exists(dirname(path)))) { 
  dir.create(dirname(path), recursive = TRUE)
  print('>> Directory created\n')}
ggsave(path, plot = plt, width = 12, height = 5)
```


# Verify data
I dounf strange data where keystoneness is NA, to test this I will look if it is a code problem or R problem.
SOLVED:
- At the moment when I calculate biomass component it has to be with before population.
```{r, eval=FALSE}
library(dplyr)
library(tidyr)

# Look at data
cols <- c("keystoneness", "dissimilarity", "prop_extinctions", "rel_pop_initial")
dat = full_summary %>% select(all_of(cols)) %>%   
  summarise(across(everything(), list(
    min  = ~min(.x, na.rm = TRUE),
    max  = ~max(.x, na.rm = TRUE),
    NAs  = ~sum(is.na(.x))
  )))  %>%
  pivot_longer(everything(),names_to  = c("variable", ".value"), names_sep = "_(?=[^_]+$)")
print(dat)

# Select cases where keystoneness is NA
na_cases = full_summary %>% filter(is.na(keystoneness)) %>% mutate(expected_keystoneness = dissimilarity * rel_pop_initial)
sum(is.na(na_cases$expected_keystoneness))

# Which simulations failed?
failed_sims = unique(na_cases$id)

#--------------------------
# Section: Replicate error
# Get parameters 
simulation_parameters = data.table::fread(paste0(dir, "/simulation-params.tsv")) %>% filter(id %in% failed_sims) 
fspecies =  na_cases$specie
unique(fspecies)

# Only specie 30????
source('/mnt/data/sur/users/mrivera/gLV/src-sims/FUN/generate-cascade-params.R')
source('/mnt/data/sur/users/mrivera/gLV/src-sims/FUN/solver-gLV.R')

replicate_error = function(index) {
  #--------------------------
  # Generate parameters
  row = simulation_parameters[index,]
  fspecie = fspecies[index]
  # Parameters
  params = gen_cascade_params(row)
  #-------------------------
  # Population at stable state
  out = solve_gLV(times = 1000, params)
  x_before = out[, ncol(out)]
  #-------------------------
  # Extinction of specie fspecie
  tmp_params <- list(
    x0 = params$x0[-fspecie],
    mu = params$mu[-fspecie],
    M = params$M[-fspecie, -fspecie, drop = FALSE]
  )
  new_out = solve_gLV(times = 1000, tmp_params)
  x_after = new_out[, ncol(new_out)]                 # Last column
 #------------------------------------
  # Section: Extinctions
  n = params$n - 1
  extinct_after = x_after <= 1e-6                       # NOW DIED
  extinct_before = tmp_params$x0 > 1e-6                 # WERE ALIVE
  n_extinctions <- sum(extinct_after & extinct_before)  # new extinctions
  props_extinctions = n_extinctions/n                   # proportion of extinctions
  #------------------------------------
  # Section: Bray-Curtis dissimilarity 
  # Remove species i from the original community.
  # Calculate
  x_removed = x_before[-fspecie]
  bray_curtis <- 1 - (2 * sum(pmin(x_removed, x_after))) / (sum(x_removed) + sum(x_after))
  #------------------------------------
  # Section: Keystoneness 
  props <- x_before / sum(x_before) # relative abundance
  keystoneness <- bray_curtis * (1 - props[fspecie])
  flag = is.na(keystoneness)
  return(flag)
}
index = 1

# 
```
